---
title: "Preparator and catalog owner contributions"
author: "Ethan Linck"
date: "4/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Building off Jessie's approach, let's sum preparator and catalog owner contributions to the "full" (multitaxa) blood dataset.

We start by loading libraries, set our working directory, and read in the data:
```{r, message=FALSE}
library(tidyverse, quietly = TRUE)
setwd("~/Dropbox/andean_range_limits/data/")
full <- read.csv("blood_data.csv")
```

Next, we'll count up the total number specimens each contributor has assigned to their name as 1) catalog owner and 2) preparator:

```{r}
cat.num.count <- full %>% group_by(`Cat.owner`) %>% summarise(count = length(`Cat.owner`))
prep.num.count <- full %>% group_by(`Preparator`) %>% summarise(count = length(`Preparator`))
```

We'll sort this in descending order, and merge dataframes so each row totals catalog specimens and prepared specimens for a single person:

```{r}
cat.num.count <- cat.num.count %>% arrange(desc(count)) 
prep.num.count <- prep.num.count %>% arrange(desc(count)) 
colnames(cat.num.count) <- c("name", "catalog_specimen_count")
colnames(prep.num.count) <- c("name", "prepared_specimen_count")
export <- inner_join(cat.num.count, prep.num.count, by = "name")
```

What does this file look like?

```{r}
head(export)
```

Last, we'll write it out as a .csv file for Chris: 

```{r}
write.csv(export, "~/Dropbox/andean_range_limits/data/total_blood_data_contributions.csv")
```